---
title: "CanvasXpress visualizations for Bioconductor data"
author: "Isaac Neuhaus"
package: CanvasXpressBio
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{CanvasXpress visualizations for Bioconductor data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

`CanvasXpressBio` connects the [CanvasXpress](https://canvasxpress.org)
interactive visualization library to core Bioconductor data structures. It does
not replace the CRAN `canvasXpress` package — it adds methods that take a
Bioconductor object directly and render it as a reproducible CanvasXpress
widget, carrying the object's annotations along for interactive grouping,
coloring and sorting.

This vignette builds **one** `SummarizedExperiment` and shows it four ways —
heatmap, volcano plot, Kaplan-Meier curve and paired boxplot — so the same
samples and genes run through every figure.

```{r load, message = FALSE}
library(CanvasXpressBio)
library(SummarizedExperiment)
```

## Why CanvasXpressBio?

Bioconductor already has excellent visualization packages. Static heatmaps
come from `ComplexHeatmap` and `pheatmap`; `EnhancedVolcano` draws volcano
plots; `survminer` draws Kaplan-Meier curves; and `InteractiveComplexHeatmap`
adds interactivity to a `ComplexHeatmap` — but through a Shiny application,
which needs a running R server to explore the figure.

`CanvasXpressBio` is complementary rather than a replacement, and is useful
when:

- **You want interactivity without a server.** Each call returns a
    self-contained CanvasXpress htmlwidget: hover for values, zoom, reorder
    and toggle groups all work in a plain knitted vignette, an R Markdown
    report, a Quarto page or an email — no Shiny runtime.
    `InteractiveComplexHeatmap` is the better choice inside a Shiny app;
    `CanvasXpressBio` is the better choice for a standalone, shareable HTML
    artefact.
- **You want one grammar across many figure types.** The same
    `SummarizedExperiment` drives the heatmap, volcano, survival curve and
    boxplot below through four small functions, so the `colData` / `rowData`
    annotations are reused consistently instead of being reshaped for a
    different package per plot.
- **You want the object's metadata to travel automatically.** `colData`
    becomes sample overlays and `rowData` becomes feature overlays, so cohort,
    time point and gene direction are available for interactive colouring and
    sorting without extra bookkeeping.

The rest of this vignette demonstrates these four views on a single object.

# One dataset for every figure

The study follows 24 patients in two cohorts, each measured at baseline and at
week 12, over a panel of twelve marker genes. The `colData` records the cohort,
the time point, the subject, a follow-up time with an event indicator, and
whether that subject's biomarker went up or down.

```{r dataset}
set.seed(1)

up_genes   <- c("IL6", "CXCL8", "MMP9", "TNF", "CCL2", "VEGFA")
down_genes <- c("ALB", "CDH1", "FOXO1", "PPARG", "ADIPOQ", "LEP")
genes      <- c(up_genes, down_genes)

patients   <- c(sprintf("PA%02d", 1:12), sprintf("PB%02d", 1:12))
cohort     <- rep(c("Cohort A", "Cohort B"), each = 12)
names(cohort) <- patients

# Sample table: each patient measured at Baseline and Week 12.
coldata <- data.frame(
    Patient   = rep(patients, each = 2),
    Condition = rep(c("Baseline", "Week 12"), times = length(patients)),
    stringsAsFactors = FALSE)
coldata$Cohort <- cohort[coldata$Patient]
rownames(coldata) <- paste(coldata$Patient,
    ifelse(coldata$Condition == "Baseline", "BL", "W12"),
    sep = "_")

# Expression: baseline around 8; at week 12 the up genes rise and the down
# genes fall. Cohort A responds consistently; cohort B is heterogeneous — some
# patients respond, a few reverse — so the biomarker moves both ways there.
baseline <- matrix(rnorm(length(genes) * length(patients), 8, 0.8),
    nrow = length(genes), dimnames = list(genes, patients))
effect   <- ifelse(genes %in% up_genes, 1.6, -1.6)
strength <- setNames(rep(1.0, length(patients)), patients)
strength[cohort == "Cohort B"] <-
    runif(sum(cohort == "Cohort B"), -0.7, 0.8)

expr <- matrix(0, nrow = length(genes), ncol = nrow(coldata),
    dimnames = list(genes, rownames(coldata)))
for (j in seq_len(nrow(coldata))) {
    p <- coldata$Patient[j]
    delta <- if (coldata$Condition[j] == "Week 12")
        effect * strength[[p]] + rnorm(length(genes), 0, 0.4) else 0
    expr[, j] <- baseline[, p] + delta
}

# Per-patient survival (cohort A does better) and the direction of the ALB
# change from baseline to week 12 (ALB is the biomarker shown in the boxplot).
alb_bl  <- expr["ALB", coldata$Condition == "Baseline"]
alb_w12 <- expr["ALB", coldata$Condition == "Week 12"]
updown  <- ifelse(alb_w12 >= alb_bl, "up", "down")
surv_time <- round(ifelse(cohort == "Cohort A",
    rexp(length(patients), 1 / 28),
    rexp(length(patients), 1 / 16)) + 2)
surv_status <- rbinom(length(patients), 1,
    ifelse(cohort == "Cohort A", .5, .75))

coldata$Time   <- surv_time[match(coldata$Patient, patients)]
coldata$Status <- surv_status[match(coldata$Patient, patients)]
coldata$updown <- updown[match(coldata$Patient, patients)]

se <- SummarizedExperiment(
    assays  = list(expr = expr),
    colData = S4Vectors::DataFrame(coldata),
    rowData = S4Vectors::DataFrame(
        Direction = ifelse(genes %in% up_genes, "Up", "Down"),
        row.names = genes))
se
```

# A heatmap from a SummarizedExperiment

`cxplot()` turns the assay matrix into an interactive heatmap. `colData` becomes
sample (column) annotations and `rowData` becomes feature (row) annotations, so
the cohort, time point and gene direction travel with the plot. `overlays` and
`var_overlays` draw chosen annotations as coloured tracks beside the map, and
`cluster = TRUE` clusters the rows and columns with dendrograms. These are the
same twelve genes highlighted in the volcano plot below.

```{r heatmap}
cxplot(se, title = "Marker-gene expression",
    overlays = c("Cohort", "Condition"), var_overlays = "Direction",
    cluster = TRUE)
```

Because `SingleCellExperiment` and `DESeqDataSet` both extend
`SummarizedExperiment`, the same method renders them without any extra code.

# A volcano plot from differential-expression results

`cxvolcano()` accepts any result table with a log-fold-change and a p-value
column — a `DESeq2::results()` object, a `limma::topTable()` data frame, or a
plain data frame. Here week 12 is contrasted against baseline; the twelve marker
genes from the heatmap are the labelled hits.

```{r volcano}
# Observed effect for the twelve markers, placed among a genome-wide background.
marker_fc <- rowMeans(expr[, coldata$Condition == "Week 12"]) -
    rowMeans(expr[, coldata$Condition == "Baseline"])
marker_p  <- 10^(-abs(marker_fc) * runif(length(genes), 2.5, 3.5))

n_bg   <- 1988
bg_fc  <- rnorm(n_bg, 0, 0.45)
bg_p   <- pmin(10^(-(abs(bg_fc) * runif(n_bg, 1.0, 3.0))), 1)

res <- data.frame(
    log2FoldChange = c(marker_fc, bg_fc),
    padj           = c(marker_p, bg_p),
    row.names      = c(genes, paste0("bg", seq_len(n_bg))))

cxvolcano(res, title = "Week 12 vs baseline",
    label_genes = genes)
```

# A Kaplan-Meier survival curve

`cxsurvival()` reads the follow-up time and event status straight from
`colData`. Using the baseline rows gives one record per patient, stratified by
cohort.

```{r km}
baseline_se <- se[, se$Condition == "Baseline"]
cxsurvival(baseline_se, time = "Time", status = "Status", group = "Cohort",
    title = "Survival by cohort")
```

# A paired boxplot with per-patient connections

`cxboxplot()` plots one marker across the two time points, draws each patient's
points, and joins the repeated measures with a line coloured by the direction of
change — faceted by cohort.

```{r boxplot}
# Paired t-test per cohort, shown as a label in each panel.
pA <- t.test(alb_w12[cohort == "Cohort A"], alb_bl[cohort == "Cohort A"],
    paired = TRUE)$p.value
pB <- t.test(alb_w12[cohort == "Cohort B"], alb_bl[cohort == "Cohort B"],
    paired = TRUE)$p.value
ptext <- function(p) if (p < 1e-4) "p<0.0001" else sprintf("p=%.3g", p)
ytop <- max(expr["ALB", ]) + 0.3

cxboxplot(se, feature = "ALB", group = "Condition", segregate = "Cohort",
    connect = "Patient", color = "updown",
    title = "ALB over time, per patient",
    colorKey = list(updown = list(
        up = "rgb(224,130,20)",
        down = "rgb(58,102,160)")),
    decorations = list(text = list(
        list(value = ytop, label = ptext(pA), panel = 1,
            smpCenter = TRUE, preTransformed = TRUE),
        list(value = ytop, label = ptext(pB), panel = 2,
            smpCenter = TRUE, preTransformed = TRUE))))
```

# Session info

```{r sessioninfo}
sessionInfo()
```
