---
title: "CanvasXpress visualizations for Bioconductor data"
author: "Isaac Neuhaus"
package: CanvasXpressBio
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{CanvasXpress visualizations for Bioconductor data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

`CanvasXpressBio` connects the [CanvasXpress](https://canvasxpress.org)
interactive visualization library to core Bioconductor data structures. It does
not replace the CRAN `canvasXpress` package — it adds methods that take a
Bioconductor object directly and render it as a reproducible CanvasXpress
widget, carrying the object's annotations along for interactive grouping,
coloring and sorting.

```{r load, message = FALSE}
library(CanvasXpressBio)
library(SummarizedExperiment)
```

# A heatmap from a SummarizedExperiment

`cxplot()` turns the assay matrix into an interactive heatmap. `colData` becomes
sample (column) annotations and `rowData` becomes feature (row) annotations.

```{r se}
set.seed(1)
se <- SummarizedExperiment(
    assays  = list(counts = matrix(
        rnorm(200), nrow = 20,
        dimnames = list(paste0("gene", 1:20), paste0("sample", 1:10)))),
    colData = S4Vectors::DataFrame(
        group = rep(c("control", "treated"), each = 5)),
    rowData = S4Vectors::DataFrame(
        chromosome = sample(c("chr1", "chr2"), 20, TRUE)))

cxplot(se, title = "Expression heatmap")
```

Because `SingleCellExperiment` and `DESeqDataSet` both extend
`SummarizedExperiment`, the same method renders them without any extra code.

## Focusing on the most variable features

Large matrices are easier to read when reduced to their most variable features.
Pass `n` to keep only the top `n` rows by variance:

```{r topn}
cxplot(se, n = 10, title = "Top 10 most variable features")
```

# A volcano plot from differential-expression results

`cxvolcano()` accepts any result table with a log-fold-change and a p-value
column — a `DESeq2::results()` object, a `limma::topTable()` data frame, or a
plain data frame.

```{r volcano}
res <- data.frame(
    log2FoldChange = rnorm(500),
    padj           = runif(500),
    row.names      = paste0("gene", 1:500))

cxvolcano(res, title = "Differential expression")
```

# A real dataset (optional)

If the `airway` data package is installed, the same call works on real RNA-seq
counts:

```{r airway, eval = requireNamespace("airway", quietly = TRUE)}
library(airway)
data("airway")
cxplot(airway, n = 25, title = "airway: top 25 variable genes")
```

# Session info

```{r sessioninfo}
sessionInfo()
```
